DisReserv (v1.0): Analytical displacement solution due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes

### 2020-09-18 manuscript/manuscript-reservoir.tex
subsection 3.2 until equation 25

We parameterize the reservoir as a grid of juxtaposed right rectangular prisms.
Each grid prism has a constant pressure contrast $\Delta p_{j}$; however, pressure variations among different prisms are allowed. 
To calculate the displacement field produced by the $jth$ prism
at the $i$th coordinates  $(x_i$, $y_i, z_i)$, we integrate the solution deduced for a single nucleus of strain (equation \ref{eq:nucleus-ui_total})
over its volume and obtain
\begin{equation}
{\bf{u}}_{i} (x_i,y_i,z_i) =  
\int_{z_1}^{z_2} \:\:
\int_{y_1}^{y_2} \:\: 
\int_{x_1}^{x_2} 
{\bf{u_{1}}}_{i} \: \:  dx^{\prime}_j dy^{\prime}_j dz^{\prime}_j 
\: + \:
\int_{z_1 - 2 \, z_c}^{z_2 - 2 \, z_c} \:\:
\int_{y_1}^{y_2} \:\: 
\int_{x_1}^{x_2} 
{\bf{u_{2}}}_{i} \:\:  dx^{\prime}_j dy^{\prime}_j dz^{\prime}_j \: ,
\label{eq:nucleus-ui_jth_prism}
\end{equation}
where the limits of the integrals are defined by the borders and the center 
of the $j$th prism in the following way: $x_1$ and $x_2$ are their south and north borders; $y_1$ and  $y_2$ are their  west and east borders; $z_1$ and $z_2$ are their depths to the top and bottom; and $z_c = \frac{1}{2}(z_1 + z_2)$ is its center depth. 
In equation \ref{eq:nucleus-ui_jth_prism}, the integrations are conducted with respect to the variables $(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$. 
The integral depending on ${\bf{u_{1}}}_{i}$,
the displacement field due to a single nucleus in the infinite space (equations \ref{eq:nucleus-u1} and  \ref{eq:nucleus-u1-vector}), is conducted over the volume of the $j$th prism. The integral depending on ${\bf{u_{2}}}_{i}$,
the displacement field due to the image nucleus (equations \ref{eq:nucleus-u2} and  \ref{eq:nucleus-u2-vector}), is conducted over the volume of a different prism
symmetrically positioned above the free surface and conveniently called "image prism".
Finally, the displacement field at the coordinates $x_i$, $y_i$  and $z_i$ due to the pressure change in the whole reservoir is defined as the sum of the displacements yielded by each prism with constant pressure $\Delta p_{j}$:
\begin{equation}
{\bf{\tilde{u}}}_{i} (x_i,y_i,z_i) = \sum_{j=1}^{M} 
\int_{z_1}^{z_2} \:\:
\int_{y_1}^{y_2} \:\: 
\int_{x_1}^{x_2} 
{\bf{u_{1}}}_{i} \: \:  dx^{\prime}_j dy^{\prime}_j dz^{\prime}_j 
\: + \: 
\int_{z_1}^{z_2} \:\:
\int_{y_1}^{y_2} \:\: 
\int_{x_1}^{x_2} 
{\bf{u_{2}}}_{i} \:\:  dx^{\prime}_j dy^{\prime}_j dz^{\prime}_j, 
\label{eq:nucleus-ui_M_prisms}
\end{equation}
where $M$ is the number of prisms setting up the reservoir model.

%From equation \ref{eq:nucleus-u1-vector},  we can get the   
%$\alpha-$components of the displacement vectors ${\bf{u_{1}}}_{i}$  (the integrand of the first integral of equation \ref{eq:nucleus-ui_M_prisms})
%\begin{equation}
%{u_{1}}_\alpha =  \frac{A  \: (1 + \nu)}{E} \: \Bigg[  \frac{\partial }{\partial \alpha} {\frac{1}{{R_1}_{ij}}} \Bigg] \: \Delta p_{j} \:\:\:  
%\label{eq:u1_alpha_component}
%\end{equation}
%where $\alpha$  belongs to the set of $ x-$ $y-$ and $z-$directions of the Cartesian coordinates system.

%From equation  \ref{eq:nucleus-u2-vector}, we can get the $ x-$ and $y-$components of the displacement vectors ${\bf{u_{2}}}_{i}$  (the integrand of the second integral of equation \ref{eq:nucleus-ui_M_prisms})
%\begin{equation}
%{u_{2}}_\alpha =  \frac{A  \: (1 + \nu)}{E} \: \Bigg[ 
%(3  - 4 \nu)    \: \frac{\partial }{\partial \alpha} {\frac{1}{{R_2}_{ij}}}\: 
%+ 2 \: z_{i} \:  
%\frac{\partial^{2}  }{\partial \alpha \partial z} {\frac{1}{{R_2}_{ij}}}  \: \Bigg]
%\Delta p_{j} ,
%\label{eq:u2_alpha_component}
%\end{equation}
%where  $\alpha$  belongs to the set of $ x-$ and $y-$directions, and the $z-$component of the displacement vectors ${\bf{u_{2}}}_{i}$  
%\begin{equation}
%{u_{2}}_z =  \frac{A  \: (1 + \nu)}{E} \: \Bigg[ 
%\: - (3  - 4 \nu)  \frac{\partial }{\partial z} {\frac{1}{{R_2}_{ij}}}\: 
%+ 2 \: z_{i} \:  
%\frac{\partial^{2}  }{\partial z^{2}} {\frac{1}{{R_2}_{ij}}}  \: \Bigg]
%\Delta p_{j}
%\label{eq:u2_z_component}
%\end{equation}

%Equation \ref{eq:u1_alpha_component} shows that the $\alpha-$component of the displacement vector ${\bf{u_{1}}}_{i}$  depends on the first derivative of ${\frac{1}{{R_1}_{ij}}}$  
%with respect to the variable $\alpha$.
%Conversely, the $\alpha-$component of the displacement vector ${\bf{u_{2}}}_{i}$  
%(equation \ref{eq:u2_alpha_component}) depends not only on the first derivative of 
%${\frac{1}{{R_2}_{ij}}}$  with respect to the variable $\alpha$ but also on the second derivative of ${\frac{1}{{R_2}_{ij}}}$  with respect to the variables $\alpha$ and $z$.

By substituting equations \ref{eq:nucleus-u1-vector} and \ref{eq:nucleus-u2-vector} into equation \ref{eq:nucleus-ui_M_prisms}, we obtain the components of the displacement field ${\bf{\tilde{u}}}_{i} (x_i,y_i,z_i)$ due to the pressure change in the whole reservoir, i.e.:
\begin{equation}
{\tilde{u}}_{{i}_\alpha} = \frac{A  \: (1 + \nu)}{E}  \sum_{j=1}^{M} 
\: \Delta p_{j}  \: \int\int\limits_{v_j}\int 
\frac{\partial }{\partial \alpha} {\frac{1}{{R_1}_{ij}}} \:  dv_j 
\: + \: 
(3  - 4 \nu)  \int\int\limits_{v_j}\int
\: \frac{\partial }{\partial \alpha} {\frac{1}{{R_2}_{ij}}}\:  dv_j
+ \: 2 \: z_{i} \:  \int\int\limits_{v_j}\int
\frac{\partial^{2}  }{\partial \alpha \partial z} {\frac{1}{{R_2}_{ij}}}  \:\:  dv_j \: , 
\label{eq:u_til_alpha}
\end{equation}
and
\begin{equation}
{\tilde{u}}_{{i}_z} = \frac{A  \: (1 + \nu)}{E}  \sum_{j=1}^{M} 
\: \Delta p_{j}  \: \int\int\limits_{v_j}\int 
\frac{\partial }{\partial z} {\frac{1}{{R_1}_{ij}}} \:  dv_j 
\: - \: (3  - 4 \nu)  \int\int\limits_{v_j}\int
\: \frac{\partial }{\partial z} {\frac{1}{{R_2}_{ij}}}\:  dv_j
+ \: 2 \: z_{i} \:  \int\int\limits_{v_j}\int
\frac{\partial^{2}  }{\partial z^{2}} {\frac{1}{{R_2}_{ij}}}  \:\:  dv_j \: , 
\label{eq:u_til_z}
\end{equation}
where $\alpha = x, y$, $dv_j$ is the $j$th element of volume of the $j$th prism whose volume is $v_j$.

In the right-hand side of equations \ref{eq:u_til_alpha} and \ref{eq:u_til_z}, the three integrals are equal to  quantities of the gravitational attraction produced by the $j$th prism considering that 
$\Delta p_{j}$ is the density of the $j$th prism and the constants are equivalent to the gravitational constant. 
The first integral corresponds to the $\alpha-$component of the gravitational attraction produced by the $j$th prism.
The second and third integrals correspond, respectively, to the $\alpha-$component of the gravitational attraction and to the $\alpha \:z-$component of the gravity gradient tensor produced by the $j$th image nucleus.

The similarity between the displacement fields due to a volume source in a half-space and the gravity field allows the use of closed expressions of the gravitational potential and its derivatives produced by the 3D right rectangular prism derived by \cite{Nagyetal2000} and (\citeyear{Nagyetal2002}).

@@

The horizontal displacement field at the $i$th coordinates ($x_i$, $y_i$  and $z_i$) due to the pressure change in the whole reservoir is calculated by 
\begin{equation}
{\tilde{u}}_{{i}_h} = \sqrt{ {\tilde{u}}_{{i}_x}^{2}  +  {\tilde{u}}_{{i}_y}^{2} }  
\label{eq:horizontal_displacement}
\end{equation}
where ${\tilde{u}}_{{i}_x}$ and ${\tilde{u}}_{{i}_y}$ are the $x-$ and $y-$ components of the displacement field at the $i$th coordinates given by equation \ref{eq:u_til_alpha}, with $\alpha = x $ and $y$.

>>>>

We parameterize the reservoir as a grid of juxtaposed right rectangular prisms.
Each grid prism has a constant pressure contrast $\Delta p_{j}$; however, pressure variations among different prisms are allowed. 
To calculate the displacement field produced by the $jth$ prism
at the $i$th coordinates  $(x_i$, $y_i, z_i)$, we integrate the solution deduced for a single nucleus of strain (equation \ref{eq:displacement_ui})
over its volume and obtain
\begin{equation}
{\bf{\tilde{u}}}_{ij} = 
\int_{z_1}^{z_2} \:\:
\int_{y_1}^{y_2} \:\: 
\int_{x_1}^{x_2} 
{\bf{u_{1}}}_{ij} \: \:  dx^{\prime}_j dy^{\prime}_j dz^{\prime}_j 
\: + \:
\int_{z_1 - 2 \, z_c}^{z_2 - 2 \, z_c} \:\:
\int_{y_1}^{y_2} \:\: 
\int_{x_1}^{x_2} 
{\bf{u_{2}}}_{ij} \:\:  dx^{\prime}_j dy^{\prime}_j dz^{\prime}_j \: ,
\label{eq:nucleus-ui_jth_prism}
\end{equation}
where the limits of the integrals are defined by the borders and the center 
of the $j$th prism in the following way: $x_1$ and $x_2$ are their south and north borders; $y_1$ and  $y_2$ are their  west and east borders; $z_1$ and $z_2$ are their depths to the top and bottom; and $z_c = \frac{1}{2}(z_1 + z_2)$ is its center depth. 
In equation \ref{eq:nucleus-ui_jth_prism}, the integrations are conducted with respect to the variables $(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$. 
The integral depending on ${\bf{u_{1}}}_{ij}$,
the displacement field due to a single nucleus in the infinite space (equations \ref{eq:nucleus-u1} and  \ref{eq:nucleus-u1-vector}), is conducted over the volume of the $j$th prism. The integral depending on ${\bf{u_{2}}}_{ij}$,
the displacement field due to the image nucleus (equations \ref{eq:nucleus-u2} and  \ref{eq:nucleus-u2-vector}), is conducted over the volume of a different prism
symmetrically positioned above the free surface and conveniently called "image prism".

By substituting equations \ref{eq:nucleus-u1-vector} and \ref{eq:nucleus-u2-vector} into equation \ref{eq:nucleus-ui_jth_prism}, we obtain the components of the displacement vector ${\bf{\tilde{u}}}_{ij}$ due to the pressure change in the $j$th prism, i.e.:
\begin{equation}
{\tilde{u}}_{{ij}_\alpha} = \frac{A  \: (1 + \nu)}{E} 
\: \Delta p_{j}  \: \left[ \int\int\limits_{v_j}\int 
\frac{\partial }{\partial \alpha} {\frac{1}{{R_1}_{ij}}} \:  dv_j 
\: + \: 
(3  - 4 \nu)  \int\int\limits_{v_j}\int
\: \frac{\partial }{\partial \alpha} {\frac{1}{{R_2}_{ij}}}\:  dv_j
+ \: 2 \: z_{i} \:  \int\int\limits_{v_j}\int
\frac{\partial^{2}  }{\partial \alpha \partial z} {\frac{1}{{R_2}_{ij}}}  \:\:  dv_j 
\right] \: , 
\label{eq:u_til_alpha}
\end{equation}
and
\begin{equation}
{\tilde{u}}_{{ij}_z} = \frac{A  \: (1 + \nu)}{E} 
\: \Delta p_{j}  \: \left[ \int\int\limits_{v_j}\int 
\frac{\partial }{\partial z} {\frac{1}{{R_1}_{ij}}} \:  dv_j 
\: - \: (3  - 4 \nu)  \int\int\limits_{v_j}\int
\: \frac{\partial }{\partial z} {\frac{1}{{R_2}_{ij}}}\:  dv_j
+ \: 2 \: z_{i} \:  \int\int\limits_{v_j}\int
\frac{\partial^{2}  }{\partial z^{2}} {\frac{1}{{R_2}_{ij}}}  \:\:  dv_j 
\right] \: , 
\label{eq:u_til_z}
\end{equation}
where $\alpha = x, y$ and $dv_j$ is the $j$th element of volume of the $j$th prism 
whose volume is $v_j$.
In the right-hand side of equations \ref{eq:u_til_alpha} and \ref{eq:u_til_z}, the three integrals have the same form of derivatives of the gravitational field produced by the 
$j$th prism. 
The first integral corresponds to the $\alpha-$component of the gravitational attraction 
produced by the $j$th prism.
The second and third integrals correspond, respectively, to the $\alpha-$component of the gravitational attraction and to the $\alpha \:z-$component of the gravitational 
gradient tensor produced by the $j$th image prism. 
The similarity between the displacement fields due to a volume source in a half-space and the gravitational field allows the use of closed expressions of the gravitational potential and its derivatives produced by the 3D right rectangular prism derived by \cite{Nagyetal2000} and (\citeyear{Nagyetal2002}).

PAREI AQUI

@@

% Displacement produced by the whole reservoir 
Finally, the displacement field at the point $(x_i, y_i, z_i)$ due to the pressure change in the whole reservoir is defined as the sum of the displacements 
${\bf{u}}_{ij}$ (equation \ref{eq:nucleus-ui_jth_prism}) yielded by each prism 
with constant pressure $\Delta p_{j}$:
\begin{equation}
{\bf{\tilde{u}}}_{ij} = \sum_{j=1}^{M} {\bf{u}}_{ij} \: ,
\label{eq:nucleus-ui_M_prisms}
\end{equation}
where $M$ is the number of prisms setting up the reservoir model.
The horizontal displacement field at the $i$th coordinates ($x_i$, $y_i$  and $z_i$) due to the pressure change in the whole reservoir is calculated by 
\begin{equation}
{\tilde{u}}_{{i}_h} = \sqrt{ {\tilde{u}}_{{i}_x}^{2}  +  {\tilde{u}}_{{i}_y}^{2} }  
\label{eq:horizontal_displacement}
\end{equation}
where ${\tilde{u}}_{{i}_x}$ and ${\tilde{u}}_{{i}_y}$ are the $x-$ and $y-$ components of the displacement field at the $i$th coordinates given by equation \ref{eq:u_til_alpha}, with $\alpha = x $ and $y$.
